---
title: "Generate expression matrix"
author: "Alexis Rapin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---
```{r knitr, echo=F}
knitr::opts_chunk$set(echo = TRUE)
```

In this notebook, microarray assay data are processed into expression matrices using the Array Power Tool (APT) suite. Raw microarray intensities CEL files are downloaded from GEO.

## Set environment and variables

```{r environment_setup, eval=T}

# Load required packages
suppressMessages(library(here))
suppressMessages(library(tidyverse))
suppressMessages(library(GEOquery)); packageVersion("GEOquery")
suppressMessages(library(affyio)); packageVersion("affyio")

# Set microarray run ID
run.id <- "R5652"

# Set the metadata file url based on GitLab v4 API (also specify the commit ID!)
run.metadata.url <- "https://sv-renku-git.epfl.ch/api/v4/projects/lisp%2Fbxd-fasted-knowledge-base/repository/files/data%2Fruns%2Fmicroarray_runs.txt/raw?ref=5016f653"

# Source project's library
source(here("lib/lib.R"))
```

## Retrieve microarray run `r run.id` metadata from the study knowledge base

The run metadata are stored in the study knowledge base in Renku and a private access GitLab token is required to download it using `wget`.

Required: Store a valid private access GitLab token in `/tmp/gitlab_token.txt` (e.g. ``$ echo '< gitlab_token >' > /tmp/gitlab_token.txt``)

```{r download_metadata, eval=F}
system(paste0("wget",
             " --header='PRIVATE-TOKEN:", readLines("/tmp/gitlab_token.txt"), "'",
             " '", run.metadata.url, "'",
             " -O ", here('data/microarray_runs.txt')))
```
```{r, eval=T}
read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
  filter(run_id == run.id) %>%
  gather
```

Also download the run samples metadata and their fields description
```{r download_samples_metadata, eval=F}
# Download samples metadata
sample.metadata.url <- read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
  filter(run_id == run.id) %>%
  select(run_samples_source_private)
system(paste0("wget",
             " --header='PRIVATE-TOKEN:", readLines("/tmp/gitlab_token.txt"), "'",
             " '", sample.metadata.url, "'",
             " -O ", here('data/run_samples.txt')))

# Download samples metadata fields descriptions
sample.metadata.fields.description.url <- read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
  filter(run_id == run.id) %>%
  select(run_samples_fields_description_source_private)
system(paste0("wget",
             " --header='PRIVATE-TOKEN:", readLines("/tmp/gitlab_token.txt"), "'",
             " '", sample.metadata.fields.description.url, "'",
             " -O /tmp/run_samples_fields_description.txt"))

# Filter out fields descriptions not present in samples metadata
read.table('/tmp/run_samples_fields_description.txt',
           sep = "\t", header = 1) %>%
  filter(name %in% (read.table(here('data/run_samples.txt'),
           sep = "\t", header = 1) %>%
           names %>%
  str_replace("\\.\\.", ": ") %>%
  str_replace("\\.", " "))) %>%
  write.table(here('data/run_samples_fields_description.txt'),
              sep = "\t", quote = F, row.names = F)
```

## Download microarray platform files needed to compute expression values

```{r list_platform_files_locations, eval=T}
# Show support files locations
read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
      filter(run_id == run.id) %>%
  dplyr::select(platform_name, platform_id, platform_support_files_misc) %>%
  unique %>%
  gather
```

Download manufacturer files.

```{r, download_platform_files, eval=F}
platform.id <- read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
      filter(run_id == run.id) %>%
  select(platform_id)

dir.create(here("data/platform_files"))
dir.create(here("data/platform_files", platform.id))
system(paste0("wget -r --no-parent -e robots=off -l1 -nd -A '.zip' https://media.affymetrix.com/analysis/downloads/lf/xta/MTA-1_0/ -P ", here("data/platform_files", platform.id)))
```

## Download expression data
Download CEL files from GEO.

```{r, download_data, eval=F}
# Dowload and extract raw data from GEO

read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
      filter(run_id == run.id) %>%
  select(data_source_public_accession) %>%
  {
  getGEOSuppFiles(GEO = .,
                  makeDirectory = FALSE,
                  baseDir = here("data"))

untar(tarfile = here("data", paste0(., "_RAW.tar")),
      exdir = here("data", "raw"))

file.remove(here("data", paste0(., "_RAW.tar")))
}
```

## Compute expression values

Compute expression values from CEL files using ``apt-probeset-summarize`` with the ``gc-sst-rma-sketch`` standard method (gc-correction,scale-intensities,rma-bg,quant-norm.sketch=-1.usepm=true.bioc=true,pm-only,med-polish).

This assumes that:
- CEL files were compressed with gzip (.gz)
- Platform files were compressed with zip (.zip) or gzip (.gz)

```{r, eval=F, echo=F}
platform.id <- read.table(here("data/microarray_runs.txt"), sep ="\t", header = 1) %>%
      filter(run_id == run.id) %>%
  select(platform_id)
```
```{r, compute_expression_values, eval=F}
# Retrieve, extract and return the names of the required platform files
platform.files <- get.platform.files(here("data/platform_files",
                                          platform.id),
                                     c("pgf", "clf", "bgp", "mps"))

# Retrieve CEL files
system(paste0('cd ', here(),
                  ' && git lfs pull --include="data/raw/*.CEL*"',
                  ' && gunzip -k data/raw/*.CEL.gz'))

# Create the output directory
dir.create(here("data/matrix"))

# Process run data with the apt-probeset-summarize tool
system(paste0("/opt/apt_2.11.3_linux_64_bit_x86_binaries/bin/apt-probeset-summarize",
              " -a gc-sst-rma-sketch",
              " -p ", platform.files[["pgf"]],
              " -c ", platform.files[["clf"]],
              " -b ", platform.files[["bgp"]],
              " -m ", platform.files[["mps"]],
              " -o ", here("data/matrix"),
              " ", here("data/raw/*.CEL")
              ))

writeLines("Intensities were processed using the apt-probeset-summarize script from the Array Power Tool suite (version 2.11.3) with the gc-sst-rma-sketch standard method.",
              file(here("data/matrix/gc-sst-rma-sketch.method_description.txt")))
```

## Verify that expression value matrix has corresponding metadata

Create a table of available samples in expression matrix versus metadata table.
```{r, check_matrix_metadata, eval=F}
sample.qc.table <- read.table(here("data/matrix", "gc-sst-rma-sketch.summary.txt"),
                header = 1) %>%
  select(-probeset_id) %>%
  names %>%
  {data.frame(Sample.name = .)} %>%
  mutate(data = "available") %>%
  full_join(read.table(here("data/run_samples.txt"),
                              header = 1, sep = "\t") %>%
  select(Sample.name) %>%
  mutate(metadata = "available"),
  by = "Sample.name") %>%
  mutate(across(c(data, metadata),
                function(x){
                  return(factor(x, levels = c(unique(x), "not available")))
                  })) %>%
  select(data, metadata)

saveRDS(sample.qc.table, here("data/sample_qc_table.txt"))
```
```{r, eval=T, echo = F}
sample.qc.table <- readRDS(here("data/sample_qc_table.txt"))
```
```{r, eval=T, echo = T}
table(sample.qc.table)
```

## Extract dates from CEL files.

```{r, get_assay_dates, eval=F}
list.files(here("data/raw"), pattern = ".CEL", full.names = T) %>%
  {
    cel.fps <- .
    get.celfile.dates(cel.fps) %>%
      data.frame(assay_date = .,
               `CEL file` = basename(cel.fps) %>%
                 str_remove(".gz$")) %>%
      write.table(here("data/assay_dates.txt"), sep = "\t", quote = T, row.names = F)
    }
```

Summarize assay dates.

```{r, summarize_assay_dates, eval=T}
read.table(here("data/assay_dates.txt"), header = 1) %>%
  group_by(assay_date) %>%
  summarise(n_samples = n())
```

#### Environment

```{r, session_infos, eval=T}
sessionInfo()
```